---
title: "Assignment 5"
author: "Max Troeger"
date: "2025-03-28"
output: pdf_document
---

```{r setup, include=FALSE}
library(readr)
nycdata <- read_csv("/Users/maxtroeger/Documents/School/data_analytics/NYC_Citywide_Annualized_Calendar_Sales_Update_20241107.csv")
```

# 1. Derived Dataset - Staten Island
a.
For the first derived dataset I chose Staten Island.
I'm looking for the interaction among the gross square footage, total number of units, and sale price terms.
Because many sale prices are 0 or ridiculously high/low, I'll need to constrain the prices to eliminate outliers.
```{r}
statenisland <- subset(nycdata,BOROUGH=="STATEN ISLAND")
statenisland$sqft <- as.numeric(statenisland$`GROSS SQUARE FEET`)
statenisland <- subset(statenisland,sqft!=0)

quartiles <- quantile(statenisland$sqft, probs=c(.25, .75), na.rm = TRUE)
IQR <- IQR(statenisland$sqft)
Lower <- quartiles[1] - 1.5*IQR
Upper <- quartiles[2] + 1.5*IQR 
statenisland <- subset(statenisland, statenisland$sqft > Lower & statenisland$sqft < Upper)
hist(statenisland$sqft)
```
b.
c.

this is a test^[footnote hehe]

# 2. Derived Dataset - Manhattan
For the second derived dataset I chose Manhattan.
Although units in Manhattan include entries of the form `BOROUGH`=`["1","MANHATTAN"]`, choosing just the former gives a sufficiently large sample size for our purposes (~96,000 observations).
Again, I'm looking for the interaction among the gross square footage, total number of units, and sale price terms with the same comment about outliers.
```{r}
manhattan <- subset(nycdata,BOROUGH=="1")
manhattan$sqft <- as.numeric(manhattan$`GROSS SQUARE FEET`)
manhattan <- subset(manhattan,sqft!=0)

quartiles <- quantile(manhattan$sqft, probs=c(.25, .75), na.rm = TRUE)
Lower <- quartiles[1] - 1.5*IQR
Upper <- quartiles[2] + 1.5*IQR 
manhattan <- subset(manhattan, manhattan$sqft > Lower & manhattan$sqft < Upper)
hist(manhattan$sqft)
```

# 3. Conclusions